A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950

### 2022-01-11 paper.Rmd
rearranging ...

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=4,comment=NA)
library(knitr)
library(eha)
g <- function(x) (log(1 + x))^3 # x in days.
ginv <- function(y) exp(y^(1/3)) - 1
```

@@

```{r infdat3}
source("R/tbl.R")
sinf <- infdat[, c("subreg", "sex", "birthdate", "exit", "event")]
x <- head(infdat, 5)
tbl(x, caption = "A few infants in the data set.")
```

@@

```{r ggplot2, fig.cap = "BB plots with and without time transform, Umeå-Skellefteå 1895-1950.", fig.height = 5}
nbths <- NROW(infdat)
##oldpar <- par(mfrow = c(2, 1))
sur <- with(infdat, Surv(enter, exit, event))
rs <- risksets(sur)
ss <- cumsum(rs$n.events)
## Natural:
xt <- rs$risktimes
att <- c(0, 28, 183, 365)
plot(xt, ss / nbths, ylim = c(0, max(ss / nbths)), type = "l",
     axes = FALSE, col = "blue", lwd = 2, xlab = "Age (days)",
     ylab = "Cum. deaths")
abline(h = 0)
axis(2, las = 1)
axis(1, at = att, labels = att)
box()
## g-transformed:
lines(g(xt) / g(365) * 365, ss / nbths, ylim = c(0, max(ss / nbths)), type = "l",
     col = "red", lwd = 2, xlab = "Age (days)",
     main = "After g-transform", ylab = "Cum. deaths")
##abline(h = 0)
##axis(2, las = 1)
axis(3, at = g(att) / g(365) * 365, labels = att, col = "red")
box()
abline(v = 0)
abline(v = g(c(28, 183, 365)) / g(365) * 365, col = "red", lty = 3)
abline(v = c(28, 183, 365), col = "blue", lty = 3)
##
legend("bottomright", legend = c("Natural", "g-transform"), col = c("blue", "red"), lty = 1)
abline(h = (ss / nbths) [28], lty = 3 )
abline(h = (ss / nbths) [180], lty = 3 )
##par(oldpar)
```

@@

We begin by studying mortality after age 28 days on the g-scale, BP predicts that it is approximately *uniform*. 
The uniform distribution, however, is not very useful in a survival analysis context, so we will
investigate a close alternative, the *exponential * distribution. Modifying the BP plot by using successive 
divison by *risk set size* instead of sample size should result in an approximately linear graph if the 
exponential assumption is valid.

@@

and the first order approximation of this cdf is

@@

and then fit a piecewise constant hazards model with cutpoints at (0, 1, 2, 4, 8, 12, 18, 24, 30, 35, 39), see 
Figure \@ref(fig:pchneo).

```{r pchneo, fig.cap = "Neonatal hazard function with external component (dashed).", fig.height = 4}
fitneo <- pchreg(Surv(enter, exit, event) ~ 1, data = neo, cuts = c(0,  1, 2, 4, 8, 12, 18, 24, 30, 35, 39))
plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "g(days)")
abline(h = 0, v = 0)
abline(h = posthaz, col = "red", lty = 2)
```

Let us take a closer look at what happens at the end of the neonatal inerval:

```{r pchneo2, fig.cap = "Neonatal hazard function with external component (dashed), right hand tail.", fig.height = 4}
##fitneo <- pchreg(Surv(enter, exit, event) ~ 1, data = neo, cuts = c(0,  1, 2, 4, 8, 12, 18, 24, 30, 39))
plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "g(days)", xlim = c(15, 39))
abline(h = 0, v = 0)
abline(h = posthaz, col = "red", lty = 2)
```

@@

### Putting it together

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=4,comment=NA)
library(knitr)
library(eha)
g <- function(x) (log(1 + x))^3 # x in days.
dg <- function(x) (3 * (log(1 + x))^2) / (1 + x)
ginv <- function(y) exp(y^(1/3)) - 1
```

@@

If `event == FALSE`, then the infant died at age (in days) given by `exit`, otherwise
the infant survived (at least) 365 days and follow-up is lost at `exit == 365`. For the
time being, no covariates are involved. So, in Table \@ref(tab:infdat3), infants No. 1--4 survived, 
while infant No. 5 died at age 162 days.

```{r infdat3}
source("R/tbl.R")
sinf <- infdat[, c("enter", "exit", "event")]
x <- head(sinf, 4)
y <- sinf[sinf$event == 1, ][1, ]
z <- rbind(x, y)
z$id <- 1:5
z <- z[, c("id", "enter", "exit", "event")]
tbl(z, caption = "A few infants in the data set.")
```

@@

```{r ggplot2, fig.cap = "BB plots with and without time transform, Umeå-Skellefteå 1895-1950."}
nbths <- NROW(infdat)
##oldpar <- par(mfrow = c(2, 1))
sur <- with(infdat, Surv(enter, exit, event))
rs <- risksets(sur)
ss <- cumsum(rs$n.events)
## Natural:
xt <- rs$risktimes
att <- c(0, 28, 183, 365)
plot(xt, ss / nbths, ylim = c(0, max(ss / nbths)), type = "l",
     axes = FALSE, col = "blue", lwd = 2, xlab = "Age (days)",
     ylab = "Cum. deaths")
abline(h = 0)
axis(2, las = 1)
axis(1, at = att, labels = att)
box()
## g-transformed:
lines(g(xt) / g(365) * 365, ss / nbths, ylim = c(0, max(ss / nbths)), type = "l",
     col = "red", lwd = 2, xlab = "Age (days)",
     main = "After g-transform", ylab = "Cum. deaths")
##abline(h = 0)
##axis(2, las = 1)
axis(3, at = g(att) / g(365) * 365, labels = att, col = "red")
box()
abline(v = 0)
abline(v = g(c(28, 183, 365)) / g(365) * 365, col = "red", lty = 3)
abline(v = c(28, 183, 365), col = "blue", lty = 3)
##
legend("bottomright", legend = c("Natural", "g-transform"), col = c("blue", "red"), lty = 1)
abline(h = (ss / nbths) [28], lty = 3 )
abline(h = (ss / nbths) [180], lty = 3 )
##par(oldpar)
```

@@

We begin by studying mortality after age 28 days on the $g$-scale, *B-P* predicts that it is approximately *uniform*. 
The uniform distribution, however, is not very useful in a survival analysis context, so we will
investigate a close alternative, the *exponential * distribution. Modifying the *B-P* plot by using successive 
divisions by *risk set size* instead of sample size should result in an approximately linear graph if the 
exponential assumption is valid.

@@

and the first order approximation of this expression is

@@

and then fit a piecewise constant hazards model with cutpoints at $g(0, 1, 2, 4, 7, 11, 16, 22, 28)$, see 
Figure \@ref(fig:pchneo).

```{r pchneo, fig.cap = "Neonatal hazard function with external component (dashed).", fig.height = 4}
neocuts <- g(c(0, 2, 4, 7, 11, 16, 22, 28))
fitneo <- pchreg(Surv(enter, exit, event) ~ 1, data = neo, cuts = neocuts)
plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "g(days)", axes = FALSE)
axis(1)
axis(2, las = 1)
box()
abline(h = 0, v = 0)
abline(h = posthaz, col = "red", lty = 2)
```

Let us take a closer look at what happens at the end of the neonatal interval, see 
Figure \@ref(fig:pchneo2).

```{r pchneo2, fig.cap = "Neonatal hazard function with external component (dashed), right hand tail.", fig.height = 4}
##fitneo <- pchreg(Surv(enter, exit, event) ~ 1, data = neo, cuts = neocuts)
par(las = 1)
plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "g(days)", xlim = c(5, 39))
abline(h = 0, v = 0)
abline(h = posthaz, col = "red", lty = 2)
```

@@

### Putting all together

Looking at the full infant mortality interval on the $g$-scale, we have a piecewise 
constant hazards model with cutpoints at $g(0, 1, 2, 4, 7, 11, 16, 22, 28)$, see 
Figure \@ref(fig:together) (where the value close to zero is truncated).

```{r together, fig.cap = "Hazard function, infant mortality, transformed time."}
allcuts <- c(neocuts, g(365))
ginfdat <- infdat
ginfdat$exit <- g(ginfdat$exit)
fitall <- pchreg(Surv(enter, exit, event) ~ 1, data = ginfdat, cuts = allcuts)
plot(fitall, fn = "haz", col = "blue", main = "", xlab = "g(Days)", axes = FALSE,
     ylim = c(0, 0.002))
pcuts <- g(c(0, 28, 183, 365))
axis(1, at = pcuts, labels = round(pcuts))
axis(2, las = 1)
box()
abline(h = 0, v = 0)
```

Now we would like to see the content of Figure \@ref(fig:together) on the natural time 
scale. If $Y = g(T)$, we have the relation between their respective hazard functions

\begin{equation*}
h_T(t) = \frac{d}{dt} g(t) h_Y\bigl(g^{-1}(t)\bigr), \quad 0 \le t \le 365,
\end{equation*}

and with $h_Y$ shown in Figure \@ref(fig:together), we get $h_T$ in Figure 
\@ref(fig:allnorm).

```{r allnorm, fig.cap = "Hazard function, infant mortality, natural time scale."}
y <- hazards(fitall, cum = FALSE)$x
hy <- hazards(fitall, cum = FALSE)$y
te <- ginv(y)
hte <- dg(te) * hy
plot(te, hte, type = "s", col = "blue", xlab = "Days", ylab = "", axes = FALSE,
     ylim = c(0, 0.002))
axis(1, at = ginv(pcuts))
axis(2, las = 1)
box()
abline(h = 0, v = 0)
```
